Level-Set method on \texttt{p4est}

### 2015-01-15 paper/tex/applications.tex
added time evolution figures for stefan problem

We propose to apply \mohammad{what about: In this section we apply ...} our approach to the study of the phase transition of a liquid melt to a solid crystaline structure. In the case of a single component melt, and in the absence of convection, the process is dominated by diffusion and can be modeled as a Stefan problem. We decompose the computational domain $\Omega$ into \mohammad{How about: Consider a computational domain $\Omega$ decomposed into ...} two subdomains $\Omega^-$ and $\Omega^+$, separated by an interface $\Gamma$. The Stefan problem describes the evolution of the temperature $T$, decomposed into $T_s$ in the solid phase $\Omega^+$ and $T_l$ in the liquid phase $\Omega^-$, as
\begin{align}
\pd{T_l}{t} & = D_l \Delta T_l \quad \mathrm{in} ~~ \Omega^-, \label{eq:stefan_heat_equation_l} \\
\pd{T_s}{t} & = D_s \Delta T_s \quad \mathrm{in} ~~ \Omega^+. \label{eq:stefan_heat_equation_s}
\end{align}
The diffusion constants $D_l$ and $D_s$ can be discontinuous across the interface. We prescribe homogeneous Neumann boundary conditions on the edge of the computational domain, $\nabla T \cdot \mathbf{n}\vert_{\partial \Omega}=0$. At the interface between the solid and the liquid phases, the temperature is given by the Gibbs-Tompson boundary condition \cite{Alexiades;Solomon;Wilson:88:The-formation-of-a-s, Alexiades;Solomon:93:Mathematical-Modelin}
\begin{equation} \label{eq:stefan_gibbs_tompson}
T_s = T_l = T_{\Gamma} = -\epsilon_c \kappa - \epsilon_v (\mathbf{V} \cdot \mathbf{n}),
\end{equation}
where $\kappa$ is the local interface curvature, $V$ is the velocity of the interface, $\mathbf{n}$ is the outward normal to the solidification front and $\epsilon_c$ and $\epsilon_v$ are the curvature and kinetic undercooling coefficients. The interface velocity $V$ is defined from the jump in the heat flux at the interface,
\begin{equation} \label{eq:stefan_velocity}
(\mathbf{V} \cdot \mathbf{n}) = - \left[ D_l \pd{T_l}{\mathbf{n}} - D_s \pd{T_s}{\mathbf{n}} \right].
\end{equation}
We choose to use an adaptive time step with a $\text{CFL} = 5$, i.e.
\begin{equation} \label{eq:stefan_dt}
\Delta t = 5 ~ \Delta x_{min} ~ \min(1,1/\max \lVert \mathbf{V} \rVert),
\end{equation}
where $\Delta x_{min}$ is the size of the smallest cell of the forest. The general procedure to solve the Stefan problem is presented in algorithm \ref{alg:stefan}.

@@

\begin{figure}
\centering
\subfigure[]{\includegraphics[width=0.48\textwidth]{figures/Stefan_small.pdf}}
\subfigure[]{\includegraphics[width=0.48\textwidth]{figures/Stefan_large.pdf}}
\caption{Scalability of the Stefan problem for small (left) and large (right) octrees with roughly 7M and 105M grid points, respectively. The solid dashed line represents perfect scaling. As expected from the scalability analysis of the individual components, we observe excellent results, illustrating the potential \mohammad{applicability?} of our solver \mohammad{algorithms?}.}
\label{fig:stefan_scaling}
\end{figure}

\begin{table}
\centering
	\begin{tabular}{|l|c|cccccc|}
	\hline
	\multirow{2}{*}{Small Test} & $P$ & 16      & 32      & 64      & 128     & 256    & 512 \\ 	                            
	                            & $e$ & $100\%$ & $95\%$  & $90\%$  & $82\%$  & $73\%$ & $64\%$ \\
	\hline
	\multirow{2}{*}{Large Test} & $P$ & 128     & 256     & 512     & 1024    & 2048   & 4096 \\ 	                            
	                            & $e$ & $100\%$ & $98\%$  & $94\%$  & $89\%$  & $73\%$ & $61\%$ \\
	\hline
	\end{tabular}
	\caption{Parallel efficiency of the total runtime for the stefan test based on the lowest number of processes for each test.}
	\label{tab:scaling_stefan} 
\end{table}

\subsection{Numerical experiments}
\mohammad{too many transitive statements and keyword ``choose''. Can we rephrase this?}
We now present the results from a large simulation of the Stefan problem. We choose to work with a $20\times20\times20$ macro-mesh and with level-10 octrees. We choose to define the Gibbs-Tompson anisotropy undercooling coefficients \eqref{eq:stefan_gibbs_tompson} as
\begin{align*}
\epsilon_c & = \left[ \epsilon_1 \left( 1+\alpha_1 \cos(3\theta_1) \right) + \epsilon_2 \left( 1+\alpha_2 \cos(3\theta_2) \right) \right] \kappa,\\
\epsilon_v & = 0,
\end{align*}
with $\theta_1$ the angle between the normal to the interface $\mathbf{n}$ and the x-axis in the $(x,y)$ plane, $\theta_2$ the angle between $\mathbf{n}$ and the x-axis in the $(x,z)$ plane, and
\begin{align*}
\epsilon_1 & = 2~(\sin(x)+\cos(y)+2)\cdot10^{-6}, & \epsilon_2 & = 2~(\sin(x)+\cos(z)+2)\cdot10^{-6}, \\
\alpha_1 & = \frac{1}{4}(\cos(x)+\sin(y)+2), & \alpha_2 & = \frac{1}{4}(\cos(x)+\sin(z)+2).
\end{align*}
Note that these expressions are arbitrary and lead to very varied crystal shapes \mohammad{How about: ... lead to a variety of shapes?}. The computation is initialized with twenty spherical seeds of radius $0.0015$ placed randomly in the domain. We choose $D_s=D_l=1$ and set the initial temperatures $T^0_l=-0.25$ and $T^0_s=0$.

We let the simulation run with \mohammad{How about: The simulation is ran on ...} $256$ mpi processes for $6$ hours and $30$ minutes, resulting in $396$ time iterations. Visualizations of the final iteration are presented in figures \ref{fig:stefan_grid} and \ref{fig:stefan_crystals}. The final iteration of the simulation consisted of $167$M grid points when a uniform grid with the equivalent finest resolution would lead to $8.59\cdot10^{12}$ grid points, i.e. over eight trillion grid points ! Our simulation used only $0.001948\%$ \mohammad{I think if we are going to report percentage, we should trim it to 1st digit} of the number of grid points needed for the same simulation on a uniform grid. This application demonstrates the potential of our approach.

\begin{figure}[ht!]
\begin{center}
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_grid_overview.png}}
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_temperature.png}}
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_grid.png}}
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_grid_zoom.png}}
\caption{Visualization of the computational mesh (a,b,c) and the temperature field (b) for the Stefan problem simulation.} \label{fig:stefan_grid}
\end{center}
\end{figure}

@@

>>>>

In this section we apply our approach to the study of the phase transition of a liquid melt to a solid crystaline structure. In the case of a single component melt, and in the absence of convection, the process is dominated by diffusion and can be modeled as a Stefan problem. We decompose the computational domain $\Omega$ into \mohammad{How about: Consider a computational domain $\Omega$ decomposed into ...} \comment{this would be the exact phrasing of another article} two subdomains $\Omega^-$ and $\Omega^+$, separated by an interface $\Gamma$. The Stefan problem describes the evolution of the temperature $T$, decomposed into $T_s$ in the solid phase $\Omega^+$ and $T_l$ in the liquid phase $\Omega^-$, as
\begin{align}
\pd{T_l}{t} & = D_l \Delta T_l \quad \mathrm{in} ~~ \Omega^-, \label{eq:stefan_heat_equation_l} \\
\pd{T_s}{t} & = D_s \Delta T_s \quad \mathrm{in} ~~ \Omega^+. \label{eq:stefan_heat_equation_s}
\end{align}
The diffusion constants $D_l$ and $D_s$ can be discontinuous across the interface. We prescribe homogeneous Neumann boundary conditions on the edge of the computational domain, $\nabla T \cdot \mathbf{n}\vert_{\partial \Omega}=0$. At the interface between the solid and the liquid phases, the temperature is given by the Gibbs-Tompson boundary condition \cite{Alexiades;Solomon;Wilson:88:The-formation-of-a-s, Alexiades;Solomon:93:Mathematical-Modelin}
\begin{equation} \label{eq:stefan_gibbs_tompson}
T_s = T_l = T_{\Gamma} = -\epsilon_c \kappa - \epsilon_v (\mathbf{V} \cdot \mathbf{n}),
\end{equation}
where $\kappa$ is the local interface curvature, $V$ is the velocity of the interface, $\mathbf{n}$ is the outward normal to the solidification front and $\epsilon_c$ and $\epsilon_v$ are the curvature and kinetic undercooling coefficients. The interface velocity $V$ is defined from the jump in the heat flux at the interface,
\begin{equation} \label{eq:stefan_velocity}
(\mathbf{V} \cdot \mathbf{n}) = - \left[ D_l \pd{T_l}{\mathbf{n}} - D_s \pd{T_s}{\mathbf{n}} \right].
\end{equation}
We choose to use an adaptive time step with a $\text{CFL} = 5$, i.e.
\begin{equation} \label{eq:stefan_dt}
\Delta t = 5 ~ \Delta x_{min} ~ \min(1,1/\max \lVert \mathbf{V} \rVert),
\end{equation}
where $\Delta x_{min}$ is the size of the smallest cell of the forest. The general procedure to solve the Stefan problem is presented in algorithm \ref{alg:stefan}.

@@

\begin{figure}
\centering
\subfigure[]{\includegraphics[width=0.48\textwidth]{figures/Stefan_small.pdf}}
\subfigure[]{\includegraphics[width=0.48\textwidth]{figures/Stefan_large.pdf}}
\caption{Scalability of the Stefan problem for small (left) and large (right) octrees with roughly 7M and 105M grid points, respectively. The solid dashed line represents perfect scaling. As expected from the scalability analysis of the individual components, we observe excellent results, illustrating the potential of our algorithms.}
\label{fig:stefan_scaling}
\end{figure}

\begin{table}
\centering
	\begin{tabular}{|l|c|cccccc|}
	\hline
	\multirow{2}{*}{Small Test} & $P$ & 16      & 32      & 64      & 128     & 256    & 512 \\ 	                            
	                            & $e$ & $100\%$ & $95\%$  & $90\%$  & $82\%$  & $73\%$ & $64\%$ \\
	\hline
	\multirow{2}{*}{Large Test} & $P$ & 128     & 256     & 512     & 1024    & 2048   & 4096 \\ 	                            
	                            & $e$ & $100\%$ & $98\%$  & $94\%$  & $89\%$  & $73\%$ & $61\%$ \\
	\hline
	\end{tabular}
	\caption{Parallel efficiency of the total runtime for the Stefan test based on the lowest number of processes for each test. \comment{frederic, what do you think about this table ?}}
	\label{tab:scaling_stefan} 
\end{table}

\subsection{Numerical experiments}
\mohammad{too many transitive statements and keyword ``choose''. Can we rephrase this?}
We now present the results from a large simulation of the Stefan problem on a $20\times20\times20$ macro-mesh and with level-10 Octrees. We choose to define the Gibbs-Tompson anisotropy undercooling coefficients \eqref{eq:stefan_gibbs_tompson} as
\begin{align*}
\epsilon_c & = \left[ \epsilon_1 \left( 1+\alpha_1 \cos(3\theta_1) \right) + \epsilon_2 \left( 1+\alpha_2 \cos(3\theta_2) \right) \right] \kappa,\\
\epsilon_v & = 0,
\end{align*}
with $\theta_1$ the angle between the normal to the interface $\mathbf{n}$ and the x-axis in the $(x,y)$ plane, $\theta_2$ the angle between $\mathbf{n}$ and the x-axis in the $(x,z)$ plane, and
\begin{align*}
\epsilon_1 & = 2~(\sin(x)+\cos(y)+2)\cdot10^{-6}, & \epsilon_2 & = 2~(\sin(x)+\cos(z)+2)\cdot10^{-6}, \\
\alpha_1 & = \frac{1}{4}(\cos(x)+\sin(y)+2), & \alpha_2 & = \frac{1}{4}(\cos(x)+\sin(z)+2).
\end{align*}
Note that these expressions are arbitrary and lead to a variety of shapes. The computation is initialized with twenty spherical seeds of radius $0.0015$ placed randomly in the domain. We choose $D_s=D_l=1$ and set the initial temperatures $T^0_l=-0.25$ and $T^0_s=0$.

The simulation was ran on $256$ mpi processes for $6$ hours and $30$ minutes, resulting in $396$ time iterations. Visualizations of the final iteration are presented in figures \ref{fig:stefan_grid}, \ref{fig:stefan_crystals} and \ref{fig:stefan_evolution}. The final iteration of the simulation consisted of $167$M grid points when a uniform grid with the equivalent finest resolution would lead to $8.59\cdot10^{12}$ grid points, i.e. over eight trillion grid points ! Our simulation used only $0.001948\%$ \mohammad{I think if we are going to report percentage, we should trim it to 1st digit} of the number of grid points needed for the same simulation on a uniform grid. This application demonstrates the potential of our approach.

\begin{figure}[ht!]
\begin{center}
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_grid_overview.png}}
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_temperature.png}}
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_grid.png}}
\subfigure[]{\includegraphics[width=0.45\textwidth]{figures/stefan_grid_zoom.png}}
\caption{Visualization of the computational mesh (a, c, d) and the temperature field (b) for the Stefan problem simulation.} \label{fig:stefan_grid}
\end{center}
\end{figure}

@@

\begin{figure}[ht!]
\begin{center}
\subfigure[]{
\includegraphics[width=.2\textwidth]{stefan/nb2_iter19.png}
\includegraphics[width=.2\textwidth]{stefan/nb2_iter39.png}
\includegraphics[width=.2\textwidth]{stefan/nb2_iter59.png}
\includegraphics[width=.2\textwidth]{stefan/nb2_iter79.png}
}
\subfigure[]{
\includegraphics[width=.2\textwidth]{stefan/nb5_iter19.png}
\includegraphics[width=.2\textwidth]{stefan/nb5_iter39.png}
\includegraphics[width=.2\textwidth]{stefan/nb5_iter59.png}
\includegraphics[width=.2\textwidth]{stefan/nb5_iter79.png}
}
\subfigure[]{
\includegraphics[width=.2\textwidth]{stefan/nb8_iter19.png}
\includegraphics[width=.2\textwidth]{stefan/nb8_iter39.png}
\includegraphics[width=.2\textwidth]{stefan/nb8_iter59.png}
\includegraphics[width=.2\textwidth]{stefan/nb8_iter79.png}
}
\subfigure[]{
\includegraphics[width=.2\textwidth]{stefan/nb12_iter19.png}
\includegraphics[width=.2\textwidth]{stefan/nb12_iter39.png}
\includegraphics[width=.2\textwidth]{stefan/nb12_iter59.png}
\includegraphics[width=.2\textwidth]{stefan/nb12_iter79.png}
}
\caption{Time evolution of four of the crystals obtained for the Stefan problem simulation. The snapshots represent, from left to right, iterations 96, 196, 296 and 396.} \label{fig:stefan_evolution}
\end{center}
\end{figure}

### 2015-01-15 paper/tex/parallel_algorithms.tex
added time evolution figures for stefan problem

Finally, in \texttt{p4est} trees are linearized, i.e. only the leaves are explicitly stored. However, \textcolor{blue}{explicit knowledge of the hierarchal structure of tree is greatly beneficial in several algorithms, e.g.\ in search operations needed for the interpolation algorithm.} Thus, we introduce a simple reconstruction algorithm which recreates a local representation of the entire ``forest'' that is only adapted to local cells and, potentially, the ghost layer. This approach is similar to the ideas introduced in \cite{Bangerth;Burstedde;Heister;etal:11:Algorithms-and-data-} and our tests show that in a typical application they usually amount to less that $1\%$ of the entire runtime. Algorithm \ref{alg:reconstruction} illustrates how this reconstruction is performed. Given a forest and a layer of ghost cells from \texttt{p4est}, the algorithm generates a local representation of the forest by recursively refining from the root until reaching the same level and location of all leaves in \texttt{p4est} and \texttt{ghost}. Note that algorithm \ref{alg:reconstruction} does not involve any communication and is load balanced provided that the initial forest is balanced. Figure \ref{fig:reconstruction} illustrates an application of algorithm \ref{alg:reconstruction}. Note how each process has independently generated a local representation of the forest that is refined to match the same leaves as in the global forest and ghost layer.
\begin{figure}[htbp]
\begin{center}
\includegraphics[width = \textwidth]{figures/reconstruct.pdf}
\end{center}
\caption{Left: A forest refined close to an interface and partitioned among four processes as indicated by colors. Right: Each process independently recreates a local forest which is refined to match the global grid and is as coarse as possible elsewhere. Note that empty cells are fictitious, i.e. they are only required to generate the hierarchal structure and are not matched by any corresponding cell in the global forest.}
\label{fig:reconstruction}
\end{figure}
\begin{algorithm}[htbp]
\caption{$H \gets \texttt{Reconstruct (}G\texttt{)}$} %\label{euclid}
\begin{algorithmic}[1]
\State $H \gets G.\texttt{macromesh()}$
\For {$tr : G.\texttt{local\_trees()}$} \Comment{traverse local cells}
	\For {$q : tr.\texttt{cells()}$}
		\State $H.\texttt{update\_tree(} tr, q \texttt{)}$
	\EndFor
\EndFor

@@

Note that all values at the intermediate point, $\mathbf{X}^\star$, and departure point, $\mathbf{X}_d$, must be calculated via interpolation from previous the grids $G^{n}$ and $G^{n-1}$. Here we use the stabilized second-order interpolation for $\phi(\mathbf{X}_d)$ and the multi-linear interpolation for $\mathbf{V}^{n+\frac{1}{2}}(\mathbf{X}^\star)$ \cite{Min;Gibou:07:A-second-order-accur}. Although parallelization of the interpolation process on a shared-memory machine is trivial, the same cannot be said for distributed-memory machines. In fact, the parallel interpolation algorithm \ref{alg:interpolation} is probably the most important contribution of this article \comment{but hasn't it been done before ?} \textcolor{blue}{Mohammad: not on adaptive since that's what makes it complicated}. Complication arises because not all departure points will reside in the domain of current process. Moreover, due to the irregular shape of partitions, one cannot even ensure they are entirely owned by neighboring processes. At best we can only expect that their locations are bounded by a halo of width $w \le \text{CFL} \: \Delta x_{\min}$ around the local partition. Of course if one enforces $\text{CFL} \le 1$, one can ensure that the halo is bounded by the ghost layer which significantly simplifies the communication problem. This assumption, however, defeats the purpose of using a semi-Lagrangian approach in the first place since we are interested in large values of $\text{CFL}$ number.

>>>>

Finally, in \texttt{p4est} trees are linearized, i.e. only the leaves are explicitly stored. However, \textcolor{blue}{explicit knowledge of the hierarchal structure of the tree is greatly beneficial in several algorithms, e.g.\ in search operations needed for the interpolation algorithm.} Thus, we introduce a simple reconstruction algorithm which recreates a local representation of the entire ``forest'' that is only adapted to local cells and, potentially, the ghost layer. This approach is similar to the ideas introduced in \cite{Bangerth;Burstedde;Heister;etal:11:Algorithms-and-data-} and our tests show that in a typical application they usually amount to less that $1\%$ of the entire runtime. Algorithm \ref{alg:reconstruction} illustrates how this reconstruction is performed. Given a forest and a layer of ghost cells from \texttt{p4est}, the algorithm generates a local representation of the forest by recursively refining from the root until reaching the same level and location of all leaves in \texttt{p4est} and \texttt{ghost}. Note that algorithm \ref{alg:reconstruction} does not involve any communication and is load balanced provided that the initial forest is balanced. Figure \ref{fig:reconstruction} illustrates an application of algorithm \ref{alg:reconstruction}. Note how each process has independently generated a local representation of the forest that is refined to match the same leaves as in the global forest and ghost layer.
\begin{figure}[htbp]
\begin{center}
\includegraphics[width = \textwidth]{figures/reconstruct.pdf}
\end{center}
\caption{Left: A forest refined close to an interface and partitioned among four processes as indicated by colors. Right: Each process independently recreates a local forest which is refined to match the global grid and is as coarse as possible elsewhere. Note that empty cells are fictitious, i.e. they are only required to generate the hierarchal structure and are not matched by any corresponding cell in the global forest.}
\label{fig:reconstruction}
\end{figure}
\begin{algorithm}[htbp]
\caption{$H \gets \texttt{Reconstruct (}G\texttt{)}$} %\label{euclid}
\begin{algorithmic}[1]
\State $H \gets G.\texttt{macromesh()}$
\For {$tr : G.\texttt{local\_trees()}$} \Comment{traverse local cells}
	\For {$q : tr.\texttt{cells()}$}
		\State $H.\texttt{update\_tree(} tr, q \texttt{)}$
	\EndFor
\EndFor

@@

Note that all values at the intermediate point, $\mathbf{X}^\star$, and departure point, $\mathbf{X}_d$, must be calculated via interpolation from previous the grids $G^{n}$ and $G^{n-1}$. Here we use the stabilized second-order interpolation for $\phi(\mathbf{X}_d)$ and the multi-linear interpolation for $\mathbf{V}^{n+\frac{1}{2}}(\mathbf{X}^\star)$ \cite{Min;Gibou:07:A-second-order-accur}. Although parallelization of the interpolation process on a shared-memory machine is trivial, the same cannot be said for distributed-memory machines. In fact, the parallel interpolation algorithm \ref{alg:interpolation} is probably the most important contribution of this article \comment{but hasn't it been done before ?} \textcolor{blue}{Mohammad: not on adaptive since that's what makes it complicated} \comment{then maybe explain in a sentence that it's easy on uniform because it's easy to tell which process the departure point belongs to ?}. Complication arises because not all departure points will reside in the domain of current process. Moreover, due to the irregular shape of partitions, one cannot even ensure they are entirely owned by neighboring processes. At best we can only expect that their locations are bounded by a halo of width $w \le \text{CFL} \: \Delta x_{\min}$ around the local partition. Of course if one enforces $\text{CFL} \le 1$, one can ensure that the halo is bounded by the ghost layer which significantly simplifies the communication problem. This assumption, however, defeats the purpose of using a semi-Lagrangian approach in the first place since we are interested in large values of $\text{CFL}$ number.
